\Title

### 2013-02-09 synmut.tex
small change in the methods
Edits in main text and supplement. Supplement needs more work. Methods need to be somewhat more detailed.

Intrapatient HIV evolution is dominated by selection on the protein level in the
arms race with the adaptive immune system. Synonymous mutations do not have an
adaptive immunity-related phenotype and are often assumed to be neutral. Here, we show
that synonymous changes in epitope-rich regions are often deleterious but still
reach high frequencies in the viral population.
We analyze longitudinal intrapatient data from the \shankaregion~part of the
envelope gene (\env) and observe that synonymous derived alleles are rarely
fixed.
Computer models suggest that such synonymous mutations have a (Malthusian) selection
coefficient of the order of $-0.002$, and that they are brought up to high
frequency by hitchhiking on neighboring beneficial nonsynonymous alleles.
We find that synonymous mutations that disrupt base pairs in RNA stems flanking
the V loops are more likely to be lost than other synonymous changes, hinting at
a direct fitness effect of such stem-loop structures in the HIV RNA.
The patterns of fixation of nonsynonymous mutations suggest that antibody
escape mutations in \shankaregion~are only transiently beneficial, either since
the immune system is catching up or because of competition between equivalent
escapes.

@@

These escape mutations are strongly selected for their effect on the amino acid
sequence of the viral proteins. Conversely, synonymous mutations are commonly
used as approximately neutral markers in studies of viral evolution. Neutral
markers are very useful since their dynamics can be compared to
that of putatively functional sites to detect purifying or directional selection
\citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}.
In addition to maintaining protein function and avoiding the adaptive immune
recognition, however, the HIV genome has to ensure efficient processing and translation,
nuclear export, and packaging into the viral capsid: all these processes operate at the RNA
level and are sensitive to synonymous changes.
On the one hand, these processes often involve some kind of RNA secondary structure.
For example, the HIV \rev{} response element (RRE) in \env{} enhances nuclear export of
full length or partially spliced viral transcripts~\citep{fernandes_hiv-1_2012}.
Another well studied case is the interaction between viral reverse transcriptase, viral ssRNA, and the host
tRNA$^\text{Lys3}$: the latter is required for priming reverse transcription
(RT) and is bound by a pseudoknotted RNA structure in the viral 5'
untranslated region~\citep{barat_interaction_1991, paillart_vitro_2002}.
On the other hand, nucleotide-level fitness effects have been observed beyond RNA structure as
well.
In most species, it is  known that synonymous codons don't evolve completely neutrally
but that some codons are favored over others \citep{plotkin_synonymous_2011};
this is expected to be particularly important in compact viruses like HIV. 
Recent studies have shown that genetically engineered HIV strains with
altered codon usage can in some cases produce more viral protein, but in general
replicate less efficiently~\citep{ngumbela_quantitative_2008,
li_codon-usage-based_2012,keating_rich_2009}.
Codon-deoptimization has been suggested as attenuation strategy for polio and 
influenza~\citep{mueller_live_2010,coleman_virus_2008}. Purifying
selection beyond the protein sequence is therefore expected
\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}; nonetheless, positive
selection via the host adaptive immune system is restricted to changes in the
amino acid sequence.

In this paper, we characterize the dynamics of synonymous mutations in \env{}
and show that a substantial fraction of these mutations is deleterious.
We argue that synonymous mutations reach high frequencies via genetic hitchhiking
thanks to the small recombination rate of HIV~\citep{neher_recombination_2010,
batorsky_estimate_2011}.
We then
compare our observations to computational models of HIV evolution and derive
estimates for the effect synonymous mutations have on fitness.
Extending the analysis of fixation probabilities to the
nonsynonymous mutations, we show that time dependent selection or strong
competition of escape mutations inside the same epitope are necessary to explain
the observed patterns of fixation and loss.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The central quantity we investigate is the probability of fixation of a
mutation, conditional on its population frequency.  A neutral mutation
segregating at frequency $\nu$ has a probability $P_\text{fix}(\nu) = \nu$ to
spread through the population and fix; in the rest of the cases, i.e.~with
probability $1-\nu$, it goes extinct. As illustrated in the inset of \FIG{aftsyn},
this is a simple consequence of the fact that
(a) exactly one of the $N$ individuals in the current population will be
the common ancestor of the entire future population at a particular locus and
(b) this ancestor has a probability $\nu$ of carrying the mutation (assuming
the neutral mutation is not preferentially associated with genomes of high or
low fitness).
Deleterious or beneficial mutations fix less or
more often than neutral ones, respectively. \FIG{aft} shows 
the time course the frequencies of all synonymous and nonsynonymous mutations
observed \env~, C2-V5, in patient p1~\citep{shankarappa_consistent_1999},
respectively. Despite many synonymous mutations reaching high frequency, very
few fix (panel~\ref{fig:aftsyn}); in constrast, many nonsynonymous mutations fix
(panel~\ref{fig:aftnonsyn}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Synonymous polymorphisms in \env, C2-V5, are mostly deleterious}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We study the dynamics and fate of synonymous mutations more quantitatively by
analyzing data from 7 patients from
\citet{shankarappa_consistent_1999,liu_selection_2006} and 3 patients from
\citet{bunnik_autologous_2008} (see methods). The data set from
\citet{shankarappa_consistent_1999,liu_selection_2006} is restricted to the
C2-V5 region of \env, while \citet{bunnik_autologous_2008} covers the
majority of \env. Considering all mutations in a
frequency interval around $\nu_0$ at some time $t_i$, we calculate the fraction
that is found at frequency 1, at frequency 0, or at intermediate frequency at
later times $t_f$. Plotting these fixed, lost, and polymorphic fraction against
the time interval $t_f-t_i$, we see that most synonymous mutations segregate for
roughly one year and are lost much more frequently than expected (panel
\ref{fig:fixp1}). The long-time probability of fixation versus extinction of
synonymous mutations is shown as a function of the initial frequency $\nu_0$ in
panel~\ref{fig:fixp2} (red line). Restricted to the region C2-V5, we find that
$\pfix$ of synonymous mutations is far below the neutral expectation.
Outside of C2-V5, using data from \citet{bunnik_autologous_2008} only, no such
reduction in $\pfix$ is found. Restricted to the C2-V5 area, the data from
\citet{bunnik_autologous_2008} is fully compatible with data from
\citet{shankarappa_consistent_1999}. The nonsynonymous seem to follow more or
less the neutral expectation (blue line) -- a point to which we will come back below.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_p1.pdf}
\label{fig:aftsyn}}\\
 \subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p1.pdf}
\label{fig:aftnonsyn}}
\caption{Time series of allele frequencies in \env, C2-V5, from
patient 1~\cite{shankarappa_consistent_1999},
divided by synonymous (panel A) and nonsynonymous (panel B) derived alleles.
While nonsynonymous mutations frequently fix, very few synonymous
mutations do even though they are frequently observed at intermediate
frequencies. Colors indicate the position of the site along the C2-V5 region
(blue to red). Inset: the fixation probability $P_\text{fix}(\nu)$ of a neutral mutation
is simply the likelihood that the future common ancestor is currently carrying
it, i.e., its frequency $\nu$.}
\label{fig:aft}
\end{center}
\end{figure}

When interpreting these results for the fixation probabilities, it is important
to distinguish between random mutations and polymorphisms at a given frequency,
which have already been filtered by selection.
A polymorphism could be beneficial to the virus and on its way to fixation. In
this case, we expect that it fixes almost surely given we see it at high
frequency. If, on the other hand, the polymorphism is deleterious it must have
gotten to high frequency by chance (genetic drift or hitchhiking), and
we expect that selection will drive it out of the population again. Hence our
observations suggest that many of the synonymous polymorphisms at intermediate
frequencies in the part of \env~that includes the hypervariable regions are
deleterious, while outside this regions polymorphisms are mostly roughly
neutral. Note that this does not imply that all synonymous mutations are neutral
-- only those mutations observed at high frequencies, which have been
experiencing selection for some time already, tend to be neutral.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{Shankarappa_fix_loss_dt_times}
\label{fig:fixp1}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_probabilities}
\label{fig:fixp2}}
\caption{Fixation and loss of synonymous mutations.
Panel A) Time
course of loss and fixation of synonymous mutations observed in a frequency interval $\nu_0$. 
In each frequency interval, the  fraction of synonymous
mutations that ultimately fix is the fixation probability conditional on the
initial frequency.
Panel B) Fixation probability of derived synonymous
alleles as a function of $\nu_0$ in C2-V5 is below the neutral
expectation indicated by the diagonal line. This suppression is not
oberserved in other parts of {\it env} or for nonsynonymous mutations.
The horizontal error bars on the abscissa are bin sizes, the vertical ones the
standard deviation after 100 patient bootstraps of the data. Data from
Refs.~\cite{shankarappa_consistent_1999, bunnik_autologous_2008}.}
\label{fig:fixp}
\end{center}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Synonymous mutations in C2-V5 tend to disrupt conserved RNA stems}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
One possible {\it a priori} explanation for lack of fixation of synonymous
mutations in C2-V5 are  secondary structures in the viral RNA. It has
been suggested early on that parts of the viral genome that has the potential to
form stems is better conserved than the
remainder~\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}.

Recently, the propensity of nucleotides of the HIV genome to form base pairs has
been measured using the SHAPE assay (a biochemical reaction preferentially
altering unpaired bases)~\citep{watts_architecture_2009}. The SHAPE assay has
shown that the variable regions V1 to V5 tend to be unpaired, while the
conserved regions between those variable regions form stems. We partition all
synonymous alleles observed at intermediate frequencies above 10-15\% depending
on their final destiny (fixation or extinction). Subsequently, we align our
sequences to the reference NL4-3 strain used in
ref.~\citep{watts_architecture_2009} and assign them SHAPE reactivities. As
shown in \FIG{SHAPEA}, the reactivities of fixed
alleles (red line) are systematically larger than of alleles are lost (blue
line) (Kolmogorov-Smirnov test, $P\approx 0.002$).
In other words, alleles that are likely to break RNA helices are also more
likely to revert and finally be lost from the population. As a control, the
average over non-observed but potentially available polymorphisms lies between
the two curves (green line), as expected (because only some of these mutations
would interfere with stem-loop formation).
To test the hypothesis that mutations in C2-V5 are lost because they break stems in the
conserved streches between the variable loops, we split the synonymous mutations in the
extended V1-V5 region further into conserved and variable regions and found that
the biggest depression in fixation probability is observed in the conserved
stems, while the variable loops show little deviation from the neutral
signature, see \FIG{SHAPEB}. This is consistent with important stem structures
in conservered regions between loops.

In addition to RNA secondary structure, we have considered other possible
explanations for a fitness defect of synonymous mutations, in particular codon
usage bias (CUB). HIV is known to prefer A-rich codons over highly expressed
human codons~\citep{jenkins_extent_2003,kuyl_biased_2012}. We do not find,
however, any evidence for a contribution of CUB to the ultimate fate of
synonymous alleles consistent with the fact that CUB between HIV and human genes
is not shrinking at the macroevolutionary level \citep{kuyl_biased_2012}. 
Furthermore, CUB in the V1-C5 region is not very different from other parts of
the HIV genome, whereas the reduced fixation probability is only observed there. In
conclusion, although we cannot exclude an effect of CUB on fitness as a general
rule, we expect it to be a minor effect in our context.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[height=0.46\linewidth]{reactivities_histograms_syn}
\label{fig:SHAPEA}}
\subfloat{\includegraphics[height=0.46\linewidth]{fixation_probabilities_VnonV}\label{fig:SHAPEB}}
\caption{Permissible synonymous mutations have
high SHAPE reactivities.
Panel A) Synonymous substitions have
substantially higher SHAPE reactivities than mutations that reach frequencies above 15\%, but are
susequently lost (restricted to the vicinity of variable loops). This difference
is quantified by a cumulative histogram ($p=0.002$, Kolmogorov-Smirnov two 
sample test). Mutations that are never
observered show an intermediate distribution of SHAPE values.
Panel B) Fixation probability of synonymous mutations in C2-V5
separately for variable loops and the connecting conserved regions, that
harbor conserved RNA stems. As expected, fixation probability is lower inside
the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
bunnik_autologous_2008, liu_selection_2006}.}
\label{fig:SHAPE}
\end{center}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Deleterious mutations are brought to high frequency by hitchhiking}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
While the observation that some fraction of synonymous mutations is deleterious
is not unexpected, it seems odd that we observe them at high population
frequency -- at least in some regions of the genome. The region of \env~ in
which we observe deleterious mutations at high frequency, however, is special in
that it undergoes frequent adaptive changes to evade recognition by neutralizing
antibodies~\cite{williamson_adaptation_2003,richman_rapid_2003}. Due to the limited amount of
recombination in HIV~\cite{neher_recombination_2010,batorsky_estimate_2011},
deleterious mutations that are linked to adaptive variants can reach high
frequency. This process is known as hitchhiking \citep{smith_hitch-hiking_1974}
or genetic draft \citep{gillespie_genetic_2000,neher_genetic_2011}.

The potential for hitchhiking is already apparent from the allele frequency
trajectories in \FIG{aft}, where many mutations appear to change rapidly in
frequency as a flock. In order
to be advected to high frequency by a linked adaptive mutation, the deleterious
effect of the mutation has to be substantially smaller than the adaptive effect.
The latter was estimated to be on the order of $s_a = 0.01$ per day~\citep{neher_recombination_2010}.
The approximate magnitude of the deleterious effects can be estimated from
\FIG{fixp1}, that shows the distribution of times for synonymous
alleles to reach the fix or get lost starting from intermediate frequencies. The
typical time to loss is of the order of 500 days. If this loss is driven by the
deleterious effect of the mutation, this corresponds to deleterious effects of
roughly $s_d \sim - 0.002$ per day.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{simulations_graduallydel.pdf}
\label{fig:simfixpvar}}\\
\subfloat{\includegraphics[width=\linewidth]{simulations_syn.pdf}
\label{fig:simsfig}}
\caption{Distribution of selection coefficients on synonymous sites. Panel A)
The depression in $P_\text{fix}$ depends on the deleterious effect size 
of the synonymous alleles. This parameter also reduces synonymous
diversity, measured by the probability of a derived allele to be found at
intermediate frequencies $P_\text{interm}$ (first inset).
Panel B) To assess the parameter space that affects synonymous fixation and
diversity, we run many simulations with random parameters for deleterious effect
size, fraction of deleterious synonymous sites, average size of escape
mutations (color, blue ($10^{-2.5}$) to red ($10^{-1.5}$)), and rate of
introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
generation). Mostly simulations with small deleterious effects and relatively
high deleterious fractions of synonymous mutations reproduce synonymous HIV-like
fixation probability and diversity. Parameters are chosen uniformly on a
logarithmic scale from the indicated windows. This simulation assumes 6
roughly equivalent escape mutations per epitope; see supplement for alternative
models.}
\label{fig:simheat}
\end{center}
\end{figure}

To get a better idea of the range of parameters that are compatible with the
observations and our interpretation, we perform computer simulations of
evolving viral populations assuming a mix of positive and purifying selection
and rare recombination.
For this purpose, we use the simulation package FFPopSim, which includes a
module dedicated to intrapatient HIV evolution~\citep{zanini_ffpopsim:_2012}.

>>>>

Intrapatient HIV evolution is dominated by selection on the protein level in the
arms race with the adaptive immune system. Synonymous mutations do not affect 
the interaction with the adaptive immune system and are often
assumed to be neutral.
We analyze longitudinal intrapatient data from the \shankaregion~part of the
envelope gene (\env) and observe that synonymous derived alleles  rarely
fix even though they often reach high frequencies in the viral population.
We find that synonymous mutations that disrupt base pairs in RNA stems flanking
the variable loops of gp120 are more likely to be lost than other synonymous
changes, hinting at a direct fitness effect of these stem-loop structures in the
HIV RNA.
Computational modeling indicates that these synonymous mutations have a
(Malthusian) selection coefficient of the order of $-0.002$, and that they are brought up to high
frequency by hitchhiking on neighboring beneficial nonsynonymous alleles.
The patterns of fixation of nonsynonymous mutations suggest that antibody
escape mutations in \shankaregion~are only transiently beneficial, either since
the immune system is catching up or because of competition between equivalent
escapes.

@@

These escape mutations are strongly selected for their effect on the amino acid
sequence of the viral proteins. Conversely, synonymous mutations are commonly
used as approximately neutral markers in studies of viral evolution. Neutral
markers are very useful since their dynamics can be compared to
that of putatively functional sites to detect purifying or directional selection
\citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}.
In addition to maintaining protein function and avoiding the adaptive immune
recognition, however, the HIV genome has to ensure efficient processing and translation,
nuclear export, and packaging into the viral capsid: all these processes operate at the RNA
level and are sensitive to synonymous changes since these processes often
depend on RNA folding.
For example, the HIV \rev{} response element (RRE) in \env{} enhances nuclear export of
full length or partially spliced viral transcripts~\citep{fernandes_hiv-1_2012}.
Another well studied case is the interaction between viral reverse transcriptase, viral ssRNA, and the host
tRNA$^\text{Lys3}$: the latter is required for priming reverse transcription
(RT) and is bound by a pseudoknotted RNA structure in the viral 5'
untranslated region~\citep{barat_interaction_1991, paillart_vitro_2002}.

Even in absense of important RNA structure synonymous codons don't
evolve completely neutrally but that some codons are favored over others in many
species \citep{plotkin_synonymous_2011}.
Recent studies have shown that genetically engineered HIV strains with
altered codon usage can in some cases produce more viral protein, but in general
replicate less efficiently~\citep{ngumbela_quantitative_2008,
li_codon-usage-based_2012,keating_rich_2009}.
Codon-deoptimization has been suggested as attenuation strategy for polio and 
influenza~\citep{mueller_live_2010,coleman_virus_2008}. Purifying
selection beyond the protein sequence is therefore expected
\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}. Positive
selection through the host's immune system, however, is restricted to
changes in the amino acid sequence.

In this paper, we characterize the dynamics of synonymous mutations in \env{}
and show that a substantial fraction of these mutations is deleterious.
We argue that synonymous mutations reach high frequencies via genetic hitchhiking
due to the small recombination rate of HIV~\citep{neher_recombination_2010,
batorsky_estimate_2011}.
We then
compare our observations to computational models of HIV evolution and derive
estimates for the effect synonymous mutations have on fitness.
Extending the analysis of fixation probabilities to the
nonsynonymous mutations, we show that time dependent selection or strong
competition of escape mutations inside the same epitope are necessary to explain
the observed patterns of fixation and loss.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The central quantity we investigate is the probability of fixation of a
mutation, conditional on its population frequency.  A neutral mutation
segregating at frequency $\nu$ has a probability $P_\text{fix}(\nu) = \nu$ to
spread through the population and fix; in the rest of the cases, i.e.~with
probability $1-\nu$, it goes extinct. As illustrated in the inset of \FIG{aftsyn},
this is a simple consequence of the fact that
(i) exactly one individual in the current population will be
the common ancestor of the entire future population at a particular locus and
(ii) this ancestor has a probability $\nu$ of carrying the mutation (assuming
the neutral mutation is not preferentially associated with genomes of high or
low fitness).
Deleterious or beneficial mutations fix less or
more often than neutral ones, respectively. \FIG{aft} shows 
the time course the frequencies of all synonymous and nonsynonymous mutations
observed \env, C2-V5, in patient p1~\citep{shankarappa_consistent_1999},
respectively. Despite many synonymous mutations reaching high frequency, 
few fix (panel~\ref{fig:aftsyn}); in constrast, many nonsynonymous mutations fix
(panel~\ref{fig:aftnonsyn}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Synonymous polymorphisms in \env, C2-V5, are mostly deleterious}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We study the dynamics and fate of synonymous mutations more quantitatively by
analyzing data from 7 patients from
\citet{shankarappa_consistent_1999,liu_selection_2006} and 3 patients from
\citet{bunnik_autologous_2008} (see methods). The data set from
\citet{shankarappa_consistent_1999,liu_selection_2006} is restricted to the
C2-V5 region of \env, while \citet{bunnik_autologous_2008} covers the
majority of \env. Considering all mutations in a
frequency interval around $\nu_0$ at some time $t_i$, we calculate the fraction
that is found at frequency 1, at frequency 0, or at intermediate frequency at
later times $t_f$. Plotting these fixed, lost, and polymorphic fraction against
the time interval $t_f-t_i$, we see that most synonymous mutations segregate for
roughly one year and are lost much more frequently than expected (panel
\ref{fig:fixp1}). The long-time probability of fixation versus extinction of
synonymous mutations is shown as a function of the initial frequency $\nu_0$ in
panel~\ref{fig:fixp2} (red line). Restricted to the region C2-V5, we find that
$\pfix$ of synonymous mutations is far below the neutral expectation.
Outside of C2-V5, using data from \citet{bunnik_autologous_2008} only, no such
reduction in $\pfix$ is found. Restricted to the C2-V5 area, the data from
\citet{bunnik_autologous_2008} is fully compatible with data from
\citet{shankarappa_consistent_1999}. The nonsynonymous mutations seem to follow
more or less the neutral expectation (blue line) -- a point to which we will come back below.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_p1.pdf}
\label{fig:aftsyn}}\\
 \subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p1.pdf}
\label{fig:aftnonsyn}}
\caption{Time series of frequencies of synonymous (A) and non-synonymous (B)
derived alleles in \env, C2-V5, from patient
1~\cite{shankarappa_consistent_1999}.
While many nonsynonymous mutations  fix, few synonymous
mutations do even though they are frequently observed at intermediate
frequencies. Colors indicate the position of the site along the C2-V5 region
(blue to red). Inset: the fixation probability $P_\text{fix}(\nu)$ of a neutral mutation
is simply the likelihood that the future common ancestor is currently carrying
it, i.e., its frequency $\nu$.}
\label{fig:aft}
\end{center}
\end{figure}

When interpreting these results for the fixation probabilities, it is important
to distinguish between random mutations and polymorphisms observed at a certain
frequency since the latter have already been filtered by selection.
A polymorphism could be beneficial to the virus and on its way to fixation. In
this case, we expect that it fixes almost surely given we see it at high
frequency. If, on the other hand, the polymorphism is deleterious it must have
gotten to high frequency by chance (genetic drift or hitchhiking), and
we expect that selection will drive it out of the population again. Hence our
observations suggest that many of the synonymous polymorphisms at intermediate
frequencies in the part of \env~that includes the hypervariable regions are
deleterious, while outside this regions polymorphisms are mostly roughly
neutral. Note that this does not imply that all synonymous mutations in this
region are neutral -- only those mutations observed at high frequencies, which
have been experiencing selection for some time already, tend to be neutral.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_times}
\label{fig:fixp1}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_probabilities}
\label{fig:fixp2}}
\caption{Fixation and loss of synonymous mutations. \comment{fix x-axis label}
Panel A) Time
course of loss and fixation of synonymous mutations observed in a frequency interval $\nu_0$. 
In each frequency interval, the  fraction of synonymous
mutations that ultimately fix is the fixation probability conditional on the
initial frequency.
Panel B) Fixation probability of derived synonymous
alleles as a function of $\nu_0$ in C2-V5 is below the neutral
expectation indicated by the diagonal line. This suppression is not
oberserved in other parts of {\it env} or for nonsynonymous mutations.
The horizontal error bars on the abscissa are bin sizes, the vertical ones the
standard deviation after 100 patient bootstraps of the data. Data from
Refs.~\cite{shankarappa_consistent_1999, bunnik_autologous_2008}.}
\label{fig:fixp}
\end{center}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Synonymous mutations in C2-V5 tend to disrupt conserved RNA stems}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
One possible {\it a priori} explanation for lack of fixation of synonymous
mutations in C2-V5 are  secondary structures in the viral RNA whose disruption
is deleterious to the virus.
It has been suggested early on that parts of the viral genome that have the
potential to form RNA stems are better conserved than the
remainder~\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}.

Recently, the propensity of nucleotides of the HIV genome to form base pairs has
been measured using the SHAPE assay (a biochemical reaction preferentially
altering unpaired bases)~\citep{watts_architecture_2009}. The SHAPE assay has
shown that the variable regions V1 to V5 tend to be unpaired, while the
conserved regions between those variable regions form stems. We partition all
synonymous alleles observed at intermediate frequencies above 10-15\% depending
on their final destiny (fixation or extinction). Subsequently, we align our
sequences to the reference NL4-3 strain used in
ref.~\citep{watts_architecture_2009} and assign them SHAPE reactivities. As
shown in \FIG{SHAPEA}, the reactivities of fixed
alleles (red histogram) are systematically larger than of alleles are lost (blue
) (Kolmogorov-Smirnov test, $P\approx 0.002$).
In other words, alleles that are likely to break RNA helices are also more
likely to revert and finally be lost from the population. The
average over all mutations that rare not observed  (green) lies between the
those that fix or get lost.
To test the hypothesis that mutations in C2-V5 are lost because they break stems in the
conserved streches between the variable loops, we consider mutations in
variable loops and conserved parts separately. The biggest depression in
fixation probability is observed in the conserved stems, while the variable
loops show little deviation from the neutral signature, see \FIG{SHAPEB}. This
is consistent with important stem structures in conservered regions between loops.

In addition to RNA secondary structure, we have considered other possible
explanations for a fitness defect of synonymous mutations, in particular codon
usage bias (CUB). HIV is known to prefer A-rich codons over highly expressed
human codons~\citep{jenkins_extent_2003,kuyl_biased_2012}. We do not find,
however, any evidence for a contribution of average CUB to the ultimate fate of
synonymous alleles consistent with that HIV does not seem to adapt its codon
usage to its human host cells at the macroevolutionary level
\citep{kuyl_biased_2012}.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[height=0.46\linewidth]{reactivities_histograms_syn}
\label{fig:SHAPEA}}
\subfloat{\includegraphics[height=0.46\linewidth]{fixation_probabilities_VnonV}\label{fig:SHAPEB}}
\caption{Permissible synonymous mutations have
high SHAPE reactivities.
Panel A) Synonymous substitions have
substantially higher SHAPE reactivities than mutations that reach frequencies above 15\%, but are
susequently lost (restricted to the region V1-V5$\pm 50bp$)????. This difference
is quantified by a cumulative histogram ($p=0.002$, Kolmogorov-Smirnov two 
sample test). Mutations that are never
observered show an intermediate distribution of SHAPE values.
Panel B) Fixation probability of synonymous mutations in C2-V5
separately for variable loops and the connecting conserved regions, that
harbor conserved RNA stems. As expected, fixation probability is lower inside
the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
bunnik_autologous_2008, liu_selection_2006}.}
\label{fig:SHAPE}
\end{center}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Deleterious mutations are brought to high frequency by hitchhiking}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
While the observation that some fraction of synonymous mutations is deleterious
is not unexpected, it seems odd that we observe them at high population
frequency -- at least in some regions of the genome. The region of \env~in
which we observe deleterious mutations at high frequency undergoes frequent
adaptive changes to evade recognition by neutralizing 
antibodies~\cite{williamson_adaptation_2003,richman_rapid_2003}. Due to the 
limited amount of recombination in HIV~\cite{neher_recombination_2010,batorsky_estimate_2011},
deleterious mutations that are linked to adaptive variants can reach high
frequency. This process is known as hitchhiking \citep{smith_hitch-hiking_1974}
or genetic draft \citep{gillespie_genetic_2000,neher_genetic_2011}.
Hitchhiking is  apparent in \FIG{aft}, that shows that many mutations
change rapidly in frequency as a flock.

The approximate magnitude of the deleterious effects can be estimated from
\FIG{fixp1}, that shows the distribution of times for synonymous
alleles to reach the fix or get lost starting from intermediate frequencies. The
typical time to loss is of the order of 500 days. If this loss is driven by the
deleterious effect of the mutation, this corresponds to deleterious effects of
roughly $s_d \sim - 0.002$ per day.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{simulations_graduallydel.pdf}
\label{fig:simfixpvar}}\\
\subfloat{\includegraphics[width=\linewidth]{simulations_syn.pdf}
\label{fig:simsfig}}
\caption{Distribution of selection coefficients on synonymous sites. Panel A)
The depression in $P_\text{fix}$ depends on the deleterious effect size 
of the synonymous alleles. This parameter also reduces synonymous
diversity, measured by the probability of a derived allele to be found at
intermediate frequencies $P_\text{interm}$ (first inset).
Panel B) To assess the parameter space that affects synonymous fixation and
diversity, we run many simulations with random parameters for deleterious effect
size, fraction of deleterious synonymous sites, average size of escape
mutations (color, blue ($10^{-2.5}$) to red ($10^{-1.5}$)), and rate of
introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
generation). Mostly simulations with small deleterious effects and relatively
high deleterious fractions of synonymous mutations reproduce synonymous HIV-like
fixation probability and diversity. Parameters are chosen uniformly on a
logarithmic scale from the indicated windows. \comment{Can't we show this
for simple epitopes? I can't seem to find a way to describe this from the
outset}.
This simulation assumes 6 roughly equivalent escape mutations per epitope; see supplement for alternative models.}
\label{fig:simheat}
\end{center}
\end{figure}

To get a better idea of the range of parameters that are compatible with the
observations and our interpretation, we performed computer simulations of
evolving viral populations assuming a mix of positive and purifying selection
and rare recombination.
For this purpose, we use the simulation package FFPopSim, which includes a
module dedicated to intrapatient HIV evolution~\citep{zanini_ffpopsim:_2012}. 
For each simulation run, we specify the deleterious effect of synonymous
mutations, the fraction of synonymous mutations that are deleterious, the
escape rate of adaptive non-synonymous mutations and the frequency of new
escapes. Note that the escape rate is sum of two factors: (i) the beneficial
effect due to the ability to evade the immune system minus (ii) the fitness
cost of the mutation in terms of structure, stability, etc. Net escape rates
in chronic infections have been estimated to be on the order of 
$s_a = 0.01$ per day~\citep{neher_recombination_2010,Asquith}.

\FIG{simfixpvar} shows simulations results for the fixation probability and the
synonymous diversity for different deleterious effects of synonymous mutations. 
We quantify synonymous diversity,$P_\text{interm}$, as the fraction of sites
with an allele at frequency $0.25 < \nu < 0.75$. The synonymous diversity
observed in patient data is indicated in the figure.
To quantify the depression of the fixation probability, we calculate
the area between the measured fixation probability and the diagonal, which is
the neutral expectation (\FIG{simfixpvar}, lower inset). If no fixation happens,
the area will be $-0.5$; if every mutation fixes, we will measure 
$A= 0.5$. In HIV, we find $A_\text{syn} \sim -0.2$  for
synonymous changes and $A_\text{nonsyn} \sim 0$ for nonsynonymous changes.

In the three simulations shown in \FIG{simfixpvar}, the fixation probability of
synonymous alleles decreases from the neutral expectation ($A_\text{syn} \sim 0$) to zero
($A_\text{syn} \sim -0.5$) as their fitness effect
worsens; the synonymous diversity plummets as well, as it is harder for
deleterious mutations to reach the frequency threshold of 0.25.
